---
title: "thor"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{thor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`thor` provides an wrapper around LMBD; the lightning memory-mapped
database.  This is an embedded key-value store; there is no server
(like SQLite) - the database exists purely on disk and uses file
locking to manage concurrent access between processes.

Key-value stores are simple systems for persistently storing values
against keys (as the name suggests).  There's really very little
extra than that!  The complications come from trying to efficiently
query the store, or patterns like "add a new value but only if the
previous value was `foo`".

This package does not provide a faithful 1:1 mapping of the
underlying LMDB C api because that requires too much care at the R
level not to crash R!  Instead, probably at the cost of some
performance, `thor` provides a set of wrappers that try to prevent
crashes by invalidating objects in the correct order.  The approach
taken in is very similar to the [python interface to LMDB;
`py-lmdb`](https://github.com/dw/py-lmdb).

Because the whole point of interacting with a database is side
effects, `thor` uses [R6](https://cran.r-project.org/package=storr)
for the interface.  This has the unfortunate effect of complicating
the documentation somewhat because R's documentation is focussed
heavily on _functions_ and the package provides only one function
(`mdb_env`) with everything else happening through *methods* of
this object, and the objects that it creates.

`thor` tries to expose the underlying LMDB interface in a nested
set of objects of increasing power (and complexity).  The objects
that the package provides are

* `mdb_env`: the environment object, which is the interface to the
  database file.  Everything starts here!

* `mdb_dbi`: a database handle.  Multiple databases may be stored
  within a single environment and if more than one is used then
  this object is passed about to control which database things
  affect.

* `mdb_txn`: a transaction object.  LMDB is a *transactional*
  database and this object is used to carry out actions within a
  transaction (such as getting and putting data).

* `mdb_cursor`: a cursor.  To go beyond basic `get`/`put`, cursors*
  *are required.  These can be used to iterate through the ##
  *database, and to find entries.

* `mdb_proxy`: a proxy for a result.  This is used to defer copying
  data from the database into R for as long as possible.  It's a
  bit of an experiment so we'll see how useful it turns out to be.

All of these objects have their own help pages, even though only
`mdb_env` has an actual function.  On those help pages every public
function described (this is the same set that is printed when
displaying the objects).  There are other functions that can be
reached using `$` - functions beginning with a `.` should be
considered **private**; using these can crash R.  Other functions
(such as `format`) exist because of the way thor uses R6.

For basic operations, one can just use the `mdb_env` object and
ignore the rest of the package.  To do more interesting things,
you'll need transactions (`mdb_txn`), and then perhaps you'll need
cursors (`mdb_cursor`).  The proxy objects are available if you use
transactions.

## The environment

The first step is to create an "environment"; this holds one or
more "databases" (though in the most simple case you can forget
that detail and just treat the environment as a database).
``` {r }
env <- thor::mdb_env(tempfile())
```

## Transactions

LMDB is _transactional_; everything that happens to the database,
read or write, happens as a transaction.  For a write transaction
either the whole transaction happens or none of it happens.  For
both read and write transactions, the "view" of the database is
consistent from the beginning to the end of a transaction.  So if
you have a read transaction and while it is doing things a write
transaction writes to the database, the read transaction does not
"see" these changes.  You can only have one write transaction at
once, but as many read transactions as you'd like.
``` {r }
txn <- env$begin(write = TRUE)
```

### Simple operations (put, get, del, etc)

To insert data into the database, use the `put` method
``` {r }
txn$put("key", "value")
```

...to get it back out again, use the `get` method
``` {r }
txn$get("key")
```

...to delete it, use the `del` method, which returns `TRUE` if
the object was deleted and `FALSE` if not
``` {r }
txn$del("key")
txn$del("key")
```

To test if an key exists or not, use the `exists` method (which
uses a cursor internally - see below)
``` {r }
txn$exists("key")

txn$put("key", "value")
txn$put("key2", "another value")
txn$put("foo", "bar")
txn$exists("key")
```

To list all keys in the database, use the `list` method
``` {r }
txn$list()
```

which takes an argument `starts_with`:
``` {r }
txn$list("k")
txn$list("f")
txn$list("x")
```

(the `exists` and `list` methods are extensions to the LMDB api)

Because the database is transactional, we can now either use
`txn$commit()` to save the changes (writing `key`, `key2` and `foo`
to the database) or `txn$abort()` to discard the changes.
``` {r }
txn$abort()
```

### Non-string data

thor (and LMDB) can handle two types of data; strings (as above)
and raw vectors.  Raw vectors can be used to serialise R objects
using `serialize`, which allows storing of arbitrary data.  This is
the approach taken by
[`redux`](https://cran.r-project.org/package=redux) amongst other
packages.

All strings can be represented in raw vectors but the reverse is
not true; character strings may not contain the null byte and the
resulting string may not make sense.  thor uses the presence of a
null byte as a heuristic when it needs to test if a value is raw or
not.

So the string "hello" can be converted to raw:
``` {r }
charToRaw("hello")
```

But the set of bytes `2a 00 ff` cannot be:
``` {r error = TRUE}
rawToChar(as.raw(c(42, 0, 255)))
```

This poses some problems for specifying and predicting return
types, which will be explored below. thor tries hard to set the
return type predictably; a few boolean arguments to the function
determine the type rather than the contents of the data.
``` {r }
txn <- env$begin(write = TRUE)
```

First, this is why one might want to store raw data in a database.
Suppose we want to store the contents of `mtcars` as a value.  It's
not a string so we can't do
``` {r error = TRUE}
txn$put("mtcars", mtcars)
```

First we should _serialise_ it to raw:
``` {r }
mtcars_ser <- serialize(mtcars, NULL)
```

which creates a fairly long string of bytes
``` {r }
str(mtcars_ser)
```

converting back from this to an R object is easy with `unserialize`
``` {r }
identical(unserialize(mtcars_ser), mtcars)

txn$put("mtcars", mtcars_ser)
txn$list()
```

When fetching the data, thor will work out that this is raw data
and return a raw vector:
``` {r }
class(txn$get("mtcars"))
```

So we can now store and retrieve arbitrary R objects into the
database.
``` {r }
identical(unserialize(txn$get("mtcars")), mtcars)
txn$del("mtcars")
```

Automatic type detection is a mixed blessing (like pitfalls with
`sapply`) and thor provides mechanisms for taming it.

Here are two values as raw vectors - one that can be converted to a
string and one that can't
``` {r }
bytes <- as.raw(c(42, 0, 255))
string <- charToRaw("hello!")

txn$put("bytes", bytes)
txn$put("string", string)
```

The value of the return type is determined both by the value of the
object and by the value of the argument `as_raw`.

| stored  | `as_raw`  | result     |
|---------|-----------|------------|
| string  | `NULL`    | character  |
| string  | `FALSE`   | character  |
| string  | `TRUE`    | raw        |
| bytes   | `NULL`    | character  |
| bytes   | `FALSE`   | error      |
| bytes   | `TRUE`    | raw        |

for example
``` {r }
txn$get("string")
```

is character because `as_raw` is `NULL` and the value _can_ be
represented as a string, while
``` {r }
txn$get("bytes")
```

is raw because the value cannot be represented as a string.
Specifying `as_raw = TRUE` will _always_ return raw because
everything can be represented as raw.  And specifying `as_raw =
FALSE` will throw an error for a value that cannot be converted
into a string.

``` {r }
txn$abort()
```

### Multi-value operations

(this whole section is an extension to the LMDB api)

``` {r }
txn <- env$begin(write = TRUE)
```

As an multi-value analog to `get`, `put` and `del`, thor implements
`mget`, `mput` and `mdel` to add, fetch and delete multiple values
at once.  The names are influenced by Redis.

``` {r }
keys <- letters[1:8]
values <- strrep(toupper(keys), 4)
```

To put these 8 values into the database at once:
``` {r }
txn$mput(keys, values)
```

All 8 values are now in the db:
``` {r }
txn$list()
```

We can fetch all values at once:
``` {r }
txn$mget(keys)
```

`mget` here always returns a list because `as_raw` is applied to
each element in turn and any one could require storing as raw.  To
force a string, vector use `as_raw = FALSE`:
``` {r }
txn$mget(keys, as_raw = FALSE)
```

Similarly, to force a raw vector, use `as_raw = TRUE`:
``` {r }
txn$mget(keys, as_raw = TRUE)
```

The `exists` method is also vectorised:
``` {r }
txn$exists(letters[1:10])
```
